Open-Source machine learning BANTER acoustic classification of beaked whale echolocation pulses

### 2023-09-14 manuscript/manuscript.qmd
Manuscript Preparation
Draft manuscript and declaration statement for submission.

For the NAtlantic dataset, the most accurate model included echolocation pulse metrics with IPI; this model had a modest increase in accuracy over the EC (only) model (EC_IPI model, Detector Model: *sampsize* = 3, *ntree* = 10,000; Event Model: *sampsize* = 3, *ntree* = 100,000). This model provided an overall correct classification rate of 97.5% for all four species (pct.correct in Confusion Matrix, @fig-natlantic a). Classification scores ranged from 75% for Gervais' beaked whales to 100% for Cuvier's and Sowerby's beaked whales. All classification results were greater than expected (Priors in @fig-natlantic a). Results from the EC_IPI model are presented here; results from the EC model were similar and can be found in supplementary materials (Supplement Fig. 1).

@@

For the SAtlantic dataset, the most accurate model considered the echolocation pulse metrics and the IPI for an event (EC_IPI Model, Detector Model: *sampsize* = 4, *ntree* = 10,000; Event Model: *sampsize* = 1, *ntree* = 100,000) and provided an overall correct classification rate of 88.7% for all five species. Classification scores ranged from a low of 50% (Sowerby's beaked whale) to a high of 100% (True's beaked whale, @fig-natlantic); all classification results were greater than expected (Priors in @fig-natlantic a). The sample sizes for the BANTER model were kept low (event sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2. Analysis with a larger sample size resulted in improved classification results for Cuvier's and Gervais' beaked whale (see Supplement Fig. 3 for EC_IPI_ALT), but resulted in the loss of Sowerby's in the final model. Results from the EC_IPI model are presented here; results from the EC and EC_IPI_ALT model can be found in supplementary materials (Supplement Fig. 2, 3).

@@

The most accurate model considered the echolocation pulses only (EC Model, Detector Model: *sampsize* = 10, *ntree* = 5,000; Event model: *sampsize* = 4, *ntree* = 10,000) and provided an overall correct classification rate of 92.3% for all four species. Classification scores ranged from a low of 86% (Cuvier's beaked whale) to 97.5% (Longman's beaked whale, @fig-hawaii). All classification results were greater than expected (Prior in @fig-hawaii a). Results from the EC model are presented here; results from EC_IPI model can be found in supplementary materials (Supplement Fig. 4).

@@

The best model considered the echolocation pulses, IPI, and environmental features (EC_IPI_ENV Model, Detector Model: *sampsize* = 1, *ntree* = 10,000; Event model: *sampsize* = 5, *ntree* = 10,000) and provided an overall correct classification rate of 91.9% for all six species. Classification scores ranged from a low of 91.2% (BW43) to 100% for Cross Seamount beaked whale (@fig-epacific a). All classification results were greater than expected (Prior in @fig-epacific a). Results from the EC_IPI_ENV model are presented here; results from the EC and EC_IPI model can be found in supplementary materials (Supplement Figs. 5, 6).

@@

While some species have distinctive differences in their pulse characteristics that can lead to strong classification despite small sample sizes (e.g. Sowerby's beaked whales in NAtlantic), other species, such as Cuvier's beaked whales, have significant variation in their pulse measurements.  For species with high variability in the predictor variables and an overlap in the range of these variables with other species, a large sample size is required to describe the true variability of these call measurements. Even with reasonably large sample sizes, the classification may suffer (e.g., SAtlantic Cuvier's beaked whale = 82.8% correct classification). We found that by increasing the *sampsize* in BANTER, we could increase the classification rate for this species in this area to 91.8% (see Supplement Fig.3). However, this resulted in the inability to include Sowerby's beaked whale in the final model. So, an increased sample size in Sowerby's would likely lead to increased classification results for other species in the model.

>>>>

For the NAtlantic dataset, the most accurate model included echolocation pulse metrics with IPI; this model had a modest increase in accuracy over the EC (only) model (EC_IPI model, Detector Model: *sampsize* = 3, *ntree* = 10,000; Event Model: *sampsize* = 3, *ntree* = 100,000). This model provided an overall correct classification rate of 97.5% for all four species (pct.correct in Confusion Matrix, @fig-natlantic a). Classification scores ranged from 75% for Gervais' beaked whales to 100% for Cuvier's and Sowerby's beaked whales. All classification results were greater than expected (Priors in @fig-natlantic a). Results from the EC_IPI model are presented here; results from the EC model were similar and can be found in supplementary materials (Supplement Fig. S1).

@@

For the SAtlantic dataset, the most accurate model considered the echolocation pulse metrics and the IPI for an event (EC_IPI Model, Detector Model: *sampsize* = 4, *ntree* = 10,000; Event Model: *sampsize* = 1, *ntree* = 100,000) and provided an overall correct classification rate of 88.7% for all five species. Classification scores ranged from a low of 50% (Sowerby's beaked whale) to a high of 100% (True's beaked whale, @fig-natlantic); all classification results were greater than expected (Priors in @fig-natlantic a). The sample sizes for the BANTER model were kept low (event sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2. Analysis with a larger sample size resulted in improved classification results for Cuvier's and Gervais' beaked whale (see Supplement Fig. S3 for EC_IPI_ALT), but resulted in the loss of Sowerby's in the final model. Results from the EC_IPI model are presented here; results from the EC and EC_IPI_ALT model can be found in supplementary materials (Supplement Fig. S2, S3).

@@

The most accurate model considered the echolocation pulses only (EC Model, Detector Model: *sampsize* = 10, *ntree* = 5,000; Event model: *sampsize* = 4, *ntree* = 10,000) and provided an overall correct classification rate of 92.3% for all four species. Classification scores ranged from a low of 86% (Cuvier's beaked whale) to 97.5% (Longman's beaked whale, @fig-hawaii). All classification results were greater than expected (Prior in @fig-hawaii a). Results from the EC model are presented here; results from EC_IPI model can be found in supplementary materials (Supplement Fig. S4).

@@

The best model considered the echolocation pulses, IPI, and environmental features (EC_IPI_ENV Model, Detector Model: *sampsize* = 1, *ntree* = 10,000; Event model: *sampsize* = 5, *ntree* = 10,000) and provided an overall correct classification rate of 91.9% for all six species. Classification scores ranged from a low of 91.2% (BW43) to 100% for Cross Seamount beaked whale (@fig-epacific a). All classification results were greater than expected (Prior in @fig-epacific a). Results from the EC_IPI_ENV model are presented here; results from the EC and EC_IPI model can be found in supplementary materials (Supplement Figs. S5, S6).

@@

While some species have distinctive differences in their pulse characteristics that can lead to strong classification despite small sample sizes (e.g. Sowerby's beaked whales in NAtlantic), other species, such as Cuvier's beaked whales, have significant variation in their pulse measurements.  For species with high variability in the predictor variables and an overlap in the range of these variables with other species, a large sample size is required to describe the true variability of these call measurements. Even with reasonably large sample sizes, the classification may suffer (e.g., SAtlantic Cuvier's beaked whale = 82.8% correct classification). We found that by increasing the *sampsize* in BANTER, we could increase the classification rate for this species in this area to 91.8% (see Supplement Fig.S3). However, this resulted in the inability to include Sowerby's beaked whale in the final model. So, an increased sample size in Sowerby's would likely lead to increased classification results for other species in the model.
